---
title: "Penguin Cleaning"
output: html_document
date: '2022-06-02'
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Starting off

In R there is "base R" which is what you get when you
you use the library() function to load additional
packages that you install.  

One way to make sure that your work is reproducible is
to keep the details well organized and to record
exactly what you do and why.  One way to help with this
is as you go on and, potentially, need to add new libraries
use the second code chunk to keep track of 
all the libraries that you use.  This makes it 
easier for other people who want to reproduce your
code -- and for yourself when you come back to the
code months later. 

Use the second chunk for this because you can
use the first chunk to set any options you want
for your whole document. 

The code chunk below shows how to load palmerpenguins.
Also load dplyr, tidyr, skimr, ggplot2, stringr.

```{r}
library(palmerpenguins)

```

You can learn more about the palmerpenguins
data by looking at this page https://allisonhorst.github.io/palmerpenguins.
The penguin package contains two data sets:
`penguins` and `penguins_raw`.  Use the View() 
function to look at the two data sets. You can also
read over the help for the two data sets (`?penguins` `?penguins_raw`  )

Our goal is to go from penguins_raw to a new 
data set penguins_new that is as close as you can get it 
to identical to penguins.   We can test if they
are the same using the code `identical(penguins, penguins_new)`. Create a code chunk at the bottom
of this file with the test code. That way you
can periodically run it.   What happens if you run
that code chunk now?

In the code chunk below create penguins_new from
the penguins_raw data set.

```{r}

```


Let's collaboratively create notes on what 
needs to be done to clean up the pengins_raw data.  

In the code chunk below use base functions to get
a sense of the data.  What are some things you notice?


```{r}

```

Another way to get a sense of a data set is to use
the `skim()` function from the skimr package.
Do that below.

```{r}

```

## Start cleaning

Now, working on cleaning the data.  Let's work both
independently and collaboratively. 

For cleaning data the tidyverse package dplyr is
extremely useful.  It makes R work similarly to 
SQL.

Here are the main "dplyr verbs." (This nice
summary is copied from Alison Horst's dplyr tutorial.) 

- `filter()`: keep rows that satisfy your conditions
- `select()`: keep or exclude some columns
- `rename()`: rename columns
- `relocate()`: move columns around
- `mutate()`: add a new column
- `group_by()` + `summarize()`: get summary statistics by group
- `across()`: apply a function across columns
- `count()`: quickly find counts for different groups
- `case_when()`: like friendly if-else

We'll also want to use various other functions for 
specific tasks.

Working below in well organized code chunks modify
your penguins_new file.  (I have started you with
a few examples.)

Sometimes you may make a mistake. But by carefully
tracking what you do you can always go back
and run all of the code, starting at the top
and get back to the point where you need to fix 
things. 

Also, dplyr is commonly used with the pipe operator.
This is really useful, but we while we are working
it is good to avoid chains that are too long. Long
pipe chains are hard to debug.

A piping example. Notice that penguins_new
in the environment doesn't change. 

```{r}
penguins_new |> 
     select(-studyName, -`Sample Number`)  
```

To make it update the penguins_new object, assign
the code back to the object. After running
the next chunk, check the data in the environment.


```{r}
penguins_new <- penguins_new |> 
                select(-studyName, -`Sample Number`) 
```
Update the above code chunk to get rid of any other
columns you don't need.  But be careful.

Now let's try another pretty simple one, rename().

Change the code with some examples. When you 
have it the way you want, assign your changes
back to the object.

```{r}
penguins_new |> rename(newname = oldname)
```

Don't worry about doing all of them at this 
point because you can always go back and make
changes.

The last example I'll include is mutate, which
you can use to create new variables or modify
existing variables.

Let's say we want to create year, what do we
need to do? How can we find a function to do 
what we want to do?

```{r}
penguins_raw |>  
     mutate(year = WHAT, sex = WHAT)
```

Now continue working on your own for a while.

After we finish this morning's workshop come back
to this and finish it up. 

As a special challenge, now add a new variable
to your data set based on data from the raw data that
is not currently included. 
